Backward Population Synthesis: Mapping the Evolutionary History of Gravitational-Wave Progenitors

### 2022-11-29 src/tex/ms.tex
Tone down figure 4 discussion.

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any; we
present inferences for the recent GWTC-3 transient catalog that suggest that the
stable mass transfer efficiency parameter may vary with primary black hole mass.  
\end{abstract}

@@

Traditionally, population synthesis studies simulate DCO populations with a
Monte-Carlo approach. Using theoretically motivated distributions of initial
parameters like age, metallicity, component mass, orbital separation, and
eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. \textbf{This method has been applied extensively to
inference on GW data to inform phenomenological models
\citep[e.g.][]{Fishbach2017, Wysocki2019, Callister2021, Delfavero2021,
Farah2022, Delfavero2022}, population synthesis for a single formation
environment \citep[e.g.][]{Belczynski2016, Stevenson2017b, Taylor2018,
Barrett2018, Mastrogiovanni2022} or contributions from multiple formation
environments \citep[e.g][]{Zevin2017, Stevenson2017, Bouffanais2019, Zevin2021,
Wong2021, Bouffanais2021, Mapelli2022}}. These approaches rely on several
assumptions, both in the simulations and model parameterizations fitted to the
GW data, which are unlikely to be fully correct. For example, there is no reason
to suppose that the entire population of DCOs should evolve with the same fixed
hyperparameters; there is no a priori reason the fitted parameterized model
should capture the relevant features of the observed DCO distributions to
correspond to the binary physics of interest; etc. A better approach would be to
compare progenitor parameters and hyperparameter settings directly to DCO
observations, treating the population synthesis model as a mapping from
progenitor to merger parameters; merger parameters can then be mapped into
observations using a gravitational waveform family, and parameter inference
proceeds in the usual way, propagating information back up the mapping chain
\citep[e.g.][]{Veitch2015}.  A similar methodology, though with \emph{fixed}
hyperparameter settings, was advanced in \citet{Andrews2018,Andrews2021}, as we
discuss in more detail below.

@@

Our results suggest that the accretion efficiency during stable mass transfer
may depend on the primary black hole mass.  In general, our method returns the
joint posterior distribution of progenitor parameters and hyperparameters for
each event, which enables a data-driven way to study the distribution of
hyperparameters.  That is, once we have a catalog of events, each has their own
posterior in the hyperparameter space, and we can employ well-known techniques
such as hierarchical Bayesian analysis to fit a population model to the
distribution of hyperparameters; this avoids making overly-specific assumptions
like fixing the hyperparameters for all types of event.

>>>>

\begin{abstract}
One promising way to extract information about stellar astrophysics from
gravitational wave catalogs is to compare the catalog to the outputs of stellar
population synthesis modeling with varying physical assumptions.  The parameter
space of physical assumptions in population synthesis is high-dimensional and
the choice of parameters that best represents the evolution of a binary system
may depend in an as-yet-to-be-determined way on the system's properties.  Here
we propose a pipeline to simultaneously infer zero-age main sequence properties
and population synthesis parameter settings controlling modeled binary evolution
from individual gravitational wave observations of merging compact binaries. Our
pipeline can efficiently explore the high-dimensional space of population
synthesis settings and progenitor system properties for each system in a catalog
of gravitational wave observations.  We apply our pipeline to observations in
the third LIGO--Virgo Gravitational-Wave Transient Catalog. We showcase the
effectiveness of this pipeline with a detailed study of the progenitor
properties and population synthesis settings that produce mergers like the
observed GW150914.  Our pipeline permits a measurement of the variation of
population synthesis parameter settings with binary properties, if any;
\textbf{we illustrate the possibility of such capability by presenting
inferences for the recent GWTC-3 transient catalog that suggest that the stable
mass transfer efficiency parameter may vary with primary black hole mass.}  
\end{abstract}

@@

Traditionally, population synthesis studies simulate DCO populations with a
Monte-Carlo approach. Using theoretically motivated distributions of initial
parameters like age, metallicity, component mass, orbital separation, and
eccentricity, initial populations are evolved with fixed astrophysical
assumptions, or population synthesis hyperparameter settings, to produce a
synthetic catalog of DCO mergers which can be compared to parameterized models
derived from observations. \textbf{This method has been applied extensively to
inference on GW data to inform phenomenological models
\citep[e.g.][]{Fishbach2017, Wysocki2019, Callister2021, Delfavero2021,
Farah2022, Delfavero2022}, population synthesis for a single formation
environment \citep[e.g.][]{Belczynski2016, Stevenson2017b, Taylor2018,
Barrett2018, Mastrogiovanni2022} or contributions from multiple formation
environments \citep[e.g][]{Zevin2017, Stevenson2017, Bouffanais2019, Zevin2021,
Wong2021, Bouffanais2021, Mapelli2022}}. These approaches rely on several
assumptions, both in the simulations and model parameterizations fitted to the
GW data, which are unlikely to be fully correct. For example, \textbf{without
knowing whether the parameterization of hyperparameters is sufficiently
accurate}, there is no reason to suppose that the entire population of DCOs
should evolve with the same fixed hyperparameters; there is no a priori reason
the fitted parameterized model should capture the relevant features of the
observed DCO distributions to correspond to the binary physics of interest; etc.
A better approach would be to compare progenitor parameters and hyperparameter
settings directly to DCO observations, treating the population synthesis model
as a mapping from progenitor to merger parameters; merger parameters can then be
mapped into observations using a gravitational waveform family, and parameter
inference proceeds in the usual way, propagating information back up the mapping
chain \citep[e.g.][]{Veitch2015}.  A similar methodology, though with
\emph{fixed} hyperparameter settings, was advanced in
\citet{Andrews2018,Andrews2021}, as we discuss in more detail below.

@@

Our results suggest that the accretion efficiency during stable mass transfer
may depend on the primary black hole mass.  In general, our method returns the
joint posterior distribution of progenitor parameters and hyperparameters for
each event, which enables a data-driven way to study the distribution of
hyperparameters.  That is, once we have a catalog of events, each has their own
posterior in the hyperparameter space, and we can employ well-known techniques
such as hierarchical Bayesian analysis to fit a population model to the
distribution of hyperparameters; this avoids making overly-specific assumptions
like fixing the hyperparameters for all types of event. \textbf{In this work we
focus on illustrating the possibililty of conducting such analysis with our
pipeline. Note that there are concerns that need to be addressed before one can
make statments about the physics, such as the number of channels considered and
specific choices of parameterization. These concerns are beyond the scope of
this paper, and we defer a detailed study of the physics related to the
population of GW events to future work.}
